\mbox{Verifying Imperative Higher-Order Programs}

### 2009-10-06 papers/esop2010/main.tex
remarques JC - 2

\begin{document}
\title{\mbox{Specifying Imperative Higher-Order Programs}}

@@

%Here, assuming we model the array in the programming language by a reference
%to a functional array, with functions {\tt length} and {\tt get} with the
%obvious meaning. Also, in the logic we use a derefence operator {\tt !!}
%instead of the usual {\tt !}.

@@

% TODO rewrite
%The {\tt iter} function and its specification raise a question: is the
%specification correct if {\tt f} modifies the array {\tt a}, which is required
%for the good behavior of {\tt iter}? The answer is clearly no; if {\tt f}
%messes up the contents of {\tt a}, {\tt iter} may not actually maintain the
%invariant.  This is an instance of the {\em aliasing problem} (see, for
%example, \cite{reynolds78popl}) and it is well known that standard Hoare logic
%does not support aliasing. In this paper we make the same choice and disallow
%aliasing in the programming language we introduce. Under this assumption, the
%specification we gave is correct. 
%
%Entirely excluding aliasing renders our system inappropriate for
%reasoning about shared mutable structures. However, it is interesting
%to see that our alias exclusion restriction, in many cases,
%corresponds to a natural condition imposed on higher-order functions
%that operate on some data structure. It is a common
%requirement~\cite{Nanevski08Awkward,Krishnaswami06} for iteration
%functions, for example, that the functional argument must not modify
%the structure to be iterated on. In our system, this fact follows from
%the restriction on effect instantiations, as will be seen in
%section~\ref{sec:syntax}. In the example, the array {\tt a}
%of the iterating function {\tt iter} must not be modified by its
%argument.

@@

\begin{figure}[tbp]
  \begin{eqnarray*}
    c(x~[\alist{κ}]) &=& \Trueml \\
    c(λx:τ.t) &=& ∀x:\ceil{τ}.c(t) \\
    c(t~t') &=& c(t) /\ c(t') \\
    c(\recml~f~(x:τ).\{p\}(t : τ',ε)\{q\}) &=&
    ∀x:\ceil{τ}.∀f:\ceil{τ->^ετ'}.  f = \ceil{v} => \\& &
    ∀s:\efft{ε}.p~s => \wpre_s(t,q~s) \\
    \wpre_s(v, q) &=& c(v) /\ q~s~\ceil{v}\\
    \wpre_s( t_{τ'->^ε τ} t' ,q) &=& \correct{t} /\ \correct{t'} /\
    fst~\ceil{t}~\ceil{t'}~s /\ \\ 
    & & ∀s':\efft{ε}∀ x:\ceil{τ}.~  snd~\ceil{t}~\ceil{t'}~s~s'~x => q~s'~x\\
    \wpre_s(\letml~x~[\alist{χ}] = v~\inml~t, q) &=&
      ∀\alist{χ}.c(v) /\ \letml~x~[\alist{χ}]=\ceil{v}~\inml~ \wpre(t,q)\\
    \wpre_s(\letml~x = ( t1 : τ,ε) ~\inml~ t2 , q) &=&
    \wpre_s( t1 ,λs':\efft{ ε }.λx:\ceil{τ}.\wpre_{s'}( t2 , q ))\\
    \wpre_s(\ifml~ t1 ~\thenml~ t2 ~\elseml~ t3 , q) &=&
      \ifml~ \ceil{ t1 }~\thenml~ \wpre_s( t2 , q))~\elseml~\wpre_s( t3 , q)) \\
      \wpre_s( (t : ε <: ε'), q) &=& \wpre_{s|_ε}(t, λs':\efft{ε}.q~(s\oplus s'))
  \end{eqnarray*}
  \caption{The weakest precondition calculus}
  \label{fig:wp}
\end{figure}

@@

As for the typing relation, we have two rules for the \letml-binding. If the
bound term is a value, we can translate the \letml-binding in the program by a
\letml-binding in the language. In the other case, the \wpre\ of $ t2 $ becomes
the argument of the \wpre\ of $ t1 $ (we denote $ε$ the effect of $ t1 $ and $
t2 $). The rule for \ifml-expressions is straightforward. The rule for
subeffecting adapts the domain of the current store and the expected domain of
the postcondition so that the types work out.

@@

let memo [e] (f0 : int -> int) (f : int ->{} int) (x : int) = 
  {pre f x /\ stores f0 !!table}
  if mem x !table then get x !table
  else
    let z = f x in
    table := set x z !table;
    z
  {r : post f x r}
\end{who}
  \caption{The code along with the specification of the {\tt memo} function}
  \label{fig:memo}
\end{figure}

Our first example is a simple memoization function {\tt memo} for (pure)
functions.  It takes as an argument a function (from {\tt int} to {\tt int}
here) and an integer and returns the same result as the functional argument
would; however, before executing the argument, it will look up the integer in
a table to see if the function hasn't been called yet on this integer. If so,
the table's value is returned, otherwise we call the function, store the
result in the table, and return it. Fig~\ref{fig:memo} shows the code along
with the specification. The specification is very simple; it states that {\tt
memo} will return a value that verifies the postcondition of {\tt f}, as long
is initially we have the precondition of f and as long as the table stores
only valid key/value pairs. As we are not interested in the exact way of the
implementation of the map, we simply model it here by a reference to a
functional map with get/set functions.

@@

TODO %TODO
\begin{figure}[tpb]
\begin{who}
  let realizes 
  let ymemo (f0 : int -> int) 
            (ff : (int ->{table} int) ->{} int ->{table} int) = 
  { forall (k : [[int ->{table} int ]]),
      { realizes f0 k} ff k { r : realizes f0 r} }
  let rec f (x : int) : int ->{table} int =
    { stores f0 !!table }
    if mem x !table then get x !table
    else
      let z = ff f x in
      table := set x z !table; z
    {r : r = f0 x /\ stores f0 !!table}
  in f
  { rf : realizes f0 rf }
\end{who}
\caption{The {\tt ymemo function} and its specification}
  \label{fig:ymemo}
\end{figure}

@@

A prototype implementation of the presented system by the name ``\who''
exists~\cite{KanigFilliatre09wml} and is freely available. It implements the
described language with type, region and effect polymorphism, higher-order
logic and store types in the logic. It takes as an input a fully annotated
program and outputs proof obligations which have to be proved to guarantee the
correctness of the program with respect to its specification. Currently, only
output to the interactive proof assistant Coq has been implemented. Our logic
is a subset of Coq's, so the translation is particularly simple.
The implementation consists of less than 3000 lines of \ocaml\ code, including
a limited form of type and effect inference.

@@

Honda, Yoshida and
Berger~\cite{Honda05anobservationally,BergerHondaYoshida05aliasing} present a
logic for imperative higher-order programs. They present their calculus using
rules in Hoare logic style, while we presented a \wpre-calculus. Their system
does not provide the same modularity than ours, as one can not reason
modularly about side effects of functional arguments. It also has the drawback
of being difficult to implement, due to the presence of a large number of
non-structural rules. Our system, starting from a correctly annotated program,
is syntax-directed and thus entirely automated. Their logic of proof
obligations is non-standard, while we obtain obligations in higher-order
logic. They deal with aliasing using equalities and disequalities between
pointers, which we chose not to do. They can reason about the classic {\em
Landin's knot}, which consists in obtaining recursion by storing a function in
a reference which calls this same reference. Our typing restriction prohibits
this particular kind of programs.  Honda, Yoshida and Berger prove a
completeness result for their system. We have proved no such result.

@@

%On the other hand, Ynot shares the
%Coq's main weakness: lack of automation. While we haven't yet presented a
%platform to automatically discharge generated proof obligations, this seems
%easily feasible, along the same vein as Régis-Gianas and
%Pottier~\cite{regis-gianas-pottier-08}, who argue that first-order SMT
%solvers~\cite{RanTin-SMTLIB} are suitable even for higher-order proof
%obligations. They also present an encoding scheme to support their claim.
%Their practical results are promising. The Ynot system can not use this kind
%of automation because it uses separation logic.  Automated provers for
%separation logic exist, but we are not aware of any attempts to use these
%tools in a higher-order setting.

%TODO also say that an approach like Ynot ou Coq doesn't give any help
%for the mechanical part of the proof.

@@

Systems for verification of first-order programs like
Why~\cite{Filliatre00a} or Spec\#~\cite{BarnettLS04} are the other
starting points of this work. These systems generally deal very well
with usual features of first-order programs, for example arrays, use
SMT solvers to discharge proof obligations and strive for the best
possible automation. We try to improve on this work by adding
higher-order features, trying to maintain the degree of automation and
ease of use.

@@

We see two possibilities to add the possibility to reason about aliasing
programs. The first one is to modify the effect system itself to allow for
aliasing. In the setting with the aliasing restriction, our regions actually
become {\em singleton regions}, regions containing exactly one memory
location. Other, more powerful effect systems allow for {\em group regions},
with the possibility to temporarily {\em focus } on a particular memory cell
in that region. We see no fundamental difficulty in adopting such an effect
system, but haven't worked out the details. A second possibility is to deal
with aliasing via equalities and disequalities over pointers, using regions to
obtain some separation automatically. On itself, we believe that this
technique does not scale, due to the explosion of needed disequalities.
However, in combination with other separation techniques, for example the
Burstall-Bornat memory model which separates automatically objects of
different structural types, this approach has worked well applied to
first-order programs.

Finally, from a programmer's point of view, it would be nice to have more
features in the programming language, such as recursive types and algebraic
data types. As long as so defined types verify a {\em positivity
condition}~\cite{paulintlca93}, these extensions would be completely
orthogonal to the weakest precondition calculus and the type and effect system
used.

>>>>

\begin{document}
\title{\mbox{Verifying Imperative Higher-Order Programs}}

@@

@@

The specification of {\tt iter} is correct, but it will be difficult to prove
in general, because it is not clear wether {\tt f} modifies the array or not,
or more precisely, if the effect variable {\tt e} does contain {\tt a} or not.
This is an instance of the {\em aliasing problem} (see, for
example, \cite{reynolds78popl}) and it is well known that standard Hoare logic
is not well-suited for programs with aliasing. Without special techniques (for
example {\em separation logic}), a large number of equalities and disequalities
are needed to reason about potentially aliased references.

To avoid this situation, we additionally introduce a simple means to exclude
programs with aliasing. The instantiation of effect variables is restricted to
the ones which do not create aliasing situations. Our system, with aliasing
exclusion enabled, is inappropriate for reasoning about shared mutable
structures. However, it is interesting to see that this alias exclusion
restriction, in many cases, corresponds to a natural condition imposed on
higher-order functions that operate on some data structure. It is a common
requirement~\cite{Nanevski08Awkward,Krishnaswami06} for iteration functions,
for example, that the functional argument must not modify the structure to be
iterated on. In our system, this fact follows from the restriction on effect
instantiations, as will be seen in section~\ref{sec:syntax}. In the example,
the array {\tt a} of the iterating function {\tt iter} must not be modified by
its argument.

@@

\begin{figure}[tbp]
  \begin{eqnarray*}
    c(x~[\alist{κ}]) &=& \Trueml \\
    c(λx:τ.t) &=& ∀x:\ceil{τ}.c(t) \\
    c(t~t') &=& c(t) /\ c(t') \\
    c(\recml~f~(x:τ).\{p\}(t : τ',ε)\{q\}) &=&
    ∀x:\ceil{τ}.∀f:\ceil{τ->^ετ'}.  f = \ceil{v} => \\& &
    ∀s:\efft{ε}.p~s => \wpre_s(t,q~s) \\
    \wpre_s(v, q) &=& c(v) /\ q~s~\ceil{v}\\
    \wpre_s( t_{τ'->^ε τ} t' ,q) &=& \correct{t} /\ \correct{t'} /\
    fst~\ceil{t}~\ceil{t'}~s /\ \\ 
    & & ∀s':\efft{ε}∀ x:\ceil{τ}.~  snd~\ceil{t}~\ceil{t'}~s~s'~x => q~s'~x\\
    \wpre_s(\letml~x~[\alist{χ}] = v~\inml~t, q) &=&
      ∀\alist{χ}.c(v) /\ \letml~x~[\alist{χ}]=\ceil{v}~\inml~ \wpre(t,q)\\
    \wpre_s(\letml~x = ( t1 : τ,ε) ~\inml~ t2 , q) &=&
    \wpre_s( t1 ,λs':\efft{ ε }.λx:\ceil{τ}.\wpre_{s'}( t2 , q ))\\
    \wpre_s(\ifml~ t1 ~\thenml~ t2 ~\elseml~ t3 , q) &=&
      \ifml~ \ceil{ t1 }~\thenml~ \wpre_s( t2 , q)~\elseml~\wpre_s( t3 , q) \\
      \wpre_s( (t : ε <: ε'), q) &=& \wpre_{s|_ε}(t, λs':\efft{ε}.q~(s\oplus s'))
  \end{eqnarray*}
  \caption{The weakest precondition calculus}
  \label{fig:wp}
\end{figure}

@@

As for the typing relation, we have two rules for the \letml-binding. If the
bound term is a value, we can translate the \letml-binding in the program by a
\letml-binding in the logic. In the other case, the \wpre\ of $ t2 $ becomes
the argument of the \wpre\ of $ t1 $ (we denote $ε$ the effect of $ t1 $ and $
t2 $). The rule for \ifml-expressions is straightforward. The rule for
subeffecting adapts the domain of the current store and the expected domain of
the postcondition so that the types work out.

@@

let memo [e] (f0 : int -> int) (x : int) = 
  {stores f0 !!table}
  if mem x !table then get x !table
  else
    let z = f0 x in
    table := set x z !table;
    z
  {r : r = f0 x /\ stores f0 !!table}
\end{who}
  \caption{The code along with the specification of the {\tt memo} function}
  \label{fig:memo}
\end{figure}

Our first example is a simple memoization function {\tt memo} for (pure)
functions.  It takes as an argument a function {\tt f} (from {\tt int} to {\tt
int} here) and an integer and returns the same result as {\tt f} would;
however, before executing the argument, it looks up the integer in a table to
see if the function hasn't been called yet on this integer. If so, the table's
value is returned, otherwise we call the function, store the result in the
table, and return it. Fig~\ref{fig:memo} shows the code along with the
specification. Since we are not interested in the implementation of the table,
we simply model it by a reference to a functional map with mem/get/set functions.

The specification of {\tt memo} is very simple. The only precondition is that
the table should only contain bindings $y |-> {\mathtt f}~y$, which is
expressend by the {\tt stores} predicate. The postcondition states that the
result is equal to {\tt f0~x} and that the table still verifies the {\tt
stores} predicate.

@@

TODO %TODO
\begin{figure}[tpb]
\begin{who}
  let realizes = .. .
  let ymemo (f0 : int -> int) 
            (ff : (int ->{table} int) ->{} int ->{table} int) = 
    { forall (k : [[int ->{table} int ]]),
        { realizes f0 k} ff k { r : realizes f0 r} }
    let rec f (x : int) =
      { stores f0 !!table }
      if mem x !table then get x !table
      else
        let z = ff f x in
        table := set x z !table; z
      {r : r = f0 x /\ stores f0 !!table}
    in f
  { rf : realizes f0 rf }
\end{who}
\caption{The {\tt ymemo function} and its specification}
  \label{fig:ymemo}
\end{figure}

@@

A prototype implementation of the presented system, called ``\who'',
 is freely available~\cite{KanigFilliatre09wml}%TODO \footnote
. It implements the
described language with type, region and effect polymorphism, higher-order
logic and store types in the logic. It takes as an input a fully annotated
program and outputs proof obligations which have to be proved to guarantee the
correctness of the program with respect to its specification. Currently, only
output to the interactive proof assistant Coq has been implemented. Our logic
is a subset of Coq's, so the translation is particularly simple.
The implementation consists of less than 3000 lines of \ocaml\ code, including
a limited form of type and effect inference.

@@

Honda, Yoshida and
Berger~\cite{Honda05anobservationally,BergerHondaYoshida05aliasing} present a
logic for imperative higher-order programs. They present their calculus using
rules in Hoare logic style, while we presented a \wpre-calculus. Their system
does not provide the same modularity than ours, as one can not reason
modularly about side effects of functional arguments. It also has the drawback
of being difficult to implement, due to the presence of a large number of
non-structural rules. Our system, starting from a correctly annotated program,
is syntax-directed and thus entirely automated. Their logic of proof
obligations is non-standard, while we obtain obligations in standard higher-order
logic.

They deal with aliasing using equalities and disequalities between
pointers, which we chose not to do.

They can reason about the classic {\em
Landin's knot}, which consists in obtaining recursion by storing a function in
a reference which calls this same reference. Our typing restriction prohibits
this particular kind of programs.  Honda, Yoshida and Berger prove a
completeness result for their system. We have proved no such result.

@@

@@

Systems for verification of first-order programs like
Why~\cite{Filliatre00a,filliatre07cav} or Spec\#~\cite{BarnettLS04} are the
other starting points of this work. These systems generally deal very well
with usual features of first-order programs, for example arrays, use SMT
solvers to discharge proof obligations and strive for the best possible
automation. We improve on this work by adding higher-order features,
while trying to maintain the degree of automation and ease of use.

@@

We see two possibilities to deal with aliasing beyond plain usage of pointer
disequalities. The first one is to modify the effect system itself to allow
more precise tracking of aliasing. In the setting with the aliasing
restriction, our regions actually become {\em singleton regions}, regions
containing exactly one memory location. Other, more powerful effect systems
allow for {\em group regions}, with the possibility to temporarily {\em focus
} on a particular memory cell in that region. We see no fundamental difficulty
in adopting such an effect system, but haven't worked out the details. A
second possibility is to deal with aliasing via equalities and disequalities
over pointers, using regions to obtain some separation automatically. On
itself, we believe that this technique does not scale, due to the explosion of
needed disequalities.  However, in combination with other separation
techniques, for example the Burstall-Bornat memory model~\cite{bornat00mpc}
which separates automatically objects of different structural types, this
approach has worked well applied to first-order programs.

Finally, from a programmer's point of view, it would be nice to have more
features in the programming language, such as recursive types and algebraic
data types. As long as defined types verify a {\em positivity
condition}~\cite{paulintlca93}, these extensions would be completely
orthogonal to the weakest precondition calculus and the type and effect system
used.
